An Algebraic Representation of Wireless Workload

### 2008-02-08 paper.tex
Work on results. Almost done...

After presenting \caps{GMM} and discussing how common concepts in traffic modeling can be expressed in precise mathematical terms within this framework, we proceed with experimental validation of \caps{GMM}. We use the technique of \textit{paired differential simulation} that the authors proposed and used in previous work~\cite{Karpinski07:realism,Karpinski07:cbr-failure}. This technique uses real wireless trace traffic as an experimental control and conducts paired experiments comparing network performance from simulations using control traffic against others using synthetically generated traffic. The results are positive: \caps{GMM} accurately reproduces network performance characteristics at all levels of the protocol stack.
%This does not solve the ``cookbook'' problem, however, because the \caps{GMM} must still be based on real traffic in order to reproduce its behavioral characteristics. It does however, reduce the problem of general traffic modeling to a more constrained problem of matrix modeling.

Our experiments with various common assumptions of uniformity and regularity in network behavior yield mixed conclusions: most straightforward assumptions of homogenous behavior distort performance results significantly, but there are notable exceptions. The clearest exception is the time domain: traffic models appear to be relatively insensitive to local perturbation of the start time of flows---so long as the overall distribution remains approximately correct. Our conclusion is that traffic modeling needs to discard its assumptions of uniformity, regularity and independence. Instead we need to move ahead trying to understand the complex dependencies between the many aspects of traffic behaviors. By reducing traffic modeling to the well-understood domain of numerical linear analysis, the General Matrix Model brings us significantly closer to the ultimate goal of having a ``cookbook'' model for generating arbitrary, realistic experimental workload patterns.

@@

The rest of the paper is organized as follows. In \Section{motivation} we discuss motivation and related work. The General Matrix Model is presented in \Section{general-matrix-model}. Our experimental and analytical methodology for evaluating traffic models is presented in \Section{methodology}, while the results of our experiments are explained and analyzed in \Section{results}. The ramifications of these results are discussed in \Section{discussion}. Finally, in \Section{conclusions}, we conclude with a discussion of how this research may be applied to current wireless studies, and how it points the way to better traffic models for the future.

\section{Motivation \& Related Work}\label{sec:motivation}\label{sec:related-work}

@@

\section{Expression of Common Properties}

@@

\section{Expression of Common Modeling Concepts}

@@

Since uniform behaviors entail homogeneity of properties with respect to each flow individually, a trace matrix can be transformed to a matrix satisfying the model's assumptions via right matrix multiplication. Right multiplication mixes the elements within each row, and if the mixing weights are uniform, the result is uniform behavior. For example, to make the roles of the nodes with respect to each flow statistically identical, we transform $S$ and $D$:
\begin{align*}
S' &= S\ones{n \times n}/n, & D' &= D\ones{n \times n}/n.
\end{align*}
This makes all the values in each flow's row identically the average of the original $n$ values. As another example, the constant bit-rate (\caps{CBR}) flow model, in its strictest sense, is an intra-flow model, which stipulates that all the packets in each flow behave identically, having identical sizes and intervals; thus each flow's bit-rate is constant, as the name implies. The strict \caps{CBR} model can be easily expressed with the packets-bytes-duration (\caps{PBD}) model used as an example in \Section{representations}. The pair of matrices $\seq{Z,V}$ is transformed to \caps{PBD} form using the vectors defined in \Section{flow-totals}: $\seq{Z,V}\Gamma\in\R^{f \times 3}$, where
\begin{align}
\Gamma=\begin{pmatrix}
\ones{} & \vec{\sigma} & \zeros{} \\
\zeros{} & \zeros{} & \vec{\tau}
\end{pmatrix} \in \R^{(d_v+d_z)\times{3}}.
\end{align}
\Figure{trace-vs-uniform} graphically depicts examples of uniform behaviors as compared to actual behaviors from traces.
%By replacing $S'$ with $S$, this can also be viewed as a set of linear constraints on the matrix $S$, providing a proscriptive definition

@@

\begin{figure}[b]
\vspace{-1em}
\begin{center}
\includegraphics[width=3.5in]{nodes-flows}%
\vspace{-0.75em}%
\caption{The number of active nodes and flows over time.} 
\label{fig:nodes-flows}
\end{center}
\vspace{-0.8em}
\end{figure}

We do not assume or claim that the traffic found at \caps{IETF}60 is representative of conference settings in general. The observed behaviors are also unlikely to resemble those found in a typical commercial or residential setting. We have chosen this trace, however, because within it can be found behaviors resembling many different types of wireless usage cases. Figure~\ref{fig:nodes-flows} shows the wide variations in the number of active flows and nodes over the course of the trace. In the night and morning hours, the traffic patterns are similar to those one might find in a moderately trafficked business or residential area. During working group sessions, we see highly concentrated, heavy usage patterns. At the zenith of activity, over 800 users, 33 thousand flows, and 1 million packets are seen in a single 10-minute trace segment. At the nadir, a lone node sent only a single 61-byte packet over the course of 10 minutes. All levels of activity between these extremes are represented. Moreover, the mix of traffic types observed changes dramatically over the course of the day, providing a wide representation of possible blends of behavior. This heterogeneity and extreme range of behaviors makes the \caps{IETF} data set ideal for this evaluation. The variety of activity gives us greater confidence that success or failure of traffic models is not tied to any specific network condition, but is broadly and generally applicable. %For a traffic model to be truly realistic, it must be realistic across many different usage cases.

Before using the traces, it is necessary to extract application-level behavior from the trace header data. First, we split the trace into individual packet flows. A flow is a series of packets sharing the following five attributes: \caps{IP} and transport protocols (raw \caps{IP}, \caps{ICMP}, \caps{TCP}, \caps{UDP}); source and destination \caps{IP} addresses and \caps{TCP}/\caps{UDP} port numbers. Next, the quantity of application-initiated data contained in each packet is calculated. For non-\caps{TCP} packets, this quantity is simply the size of the transport-layer payload, but for \caps{TCP} the calculation is more complicated: only new data transfers, explicitly initiated by the application are counted. Data retransmitted by \caps{TCP} is disregarded, and empty \caps{ACK}s are ignored. \caps{SYN} and \caps{FIN} flags in packets (even empty ones) are counted as a single byte each, since they are explicitly signaled by the application. %After processing, we have a collection of flows, each with a sequence of timestamps and the amount of application-initiated data sent at that time.

@@

We have selected six performance metrics to present here. They are commonly used as indicators of network performance at the application, network, and link layers of the protocol stack:
\begin{enumerate}
\setlength{\itemsep}{0em}
\item \textbf{Application:} average end-to-end delay, packet delivery ratio, received throughput.
\item \textbf{Network:} \caps{AODV} control overhead (\caps{RREQ/RREP/RERR}), packet drops in \caps{IP} routing queues.
\item \textbf{Link:} 802.11 control overhead (\caps{RTS/CTS/ACK}).
\end{enumerate}
These metrics are commonly used to evaluate wireless protocols. We have examined a broad variety of other wireless network performance metrics, and although we do not have room to present or discuss them here, the results shown are representative of the overall realism of the traffic models.
%For each of the 144 10-minute trace segments, we have run simulations using each of the nine traffic models, for a total of 1,296 simulations.
%Unfortunately, not all of the simulations were able to run to completion. In some scenarios, the number of nodes and flows caused Qualnet to simply fail. We were, however, able to get complete results for 72.5\% of the simulations, providing a clear view of overall trends.\!\footnote{That 27.5\% of simulations did not finish may seem like a cause for concern, but it does not in fact invalidate our results. The simulations that did not finish are those for which the number of flows and nodes overwhelmed the capacity of the simulator. Qualnet cannot simulate more than circa 650 nodes sending 35,000 flows. While it would be of great interest to examine results for these large-scale scenarios, they are essentially useless for comparison to other simulations studies: other studies do not simulate this quantity of nodes and flows precisely because even the best existing simulators do not support it. Once the simulators become more scalable and robust, our analysis can be extended to these scenarios as well.}

@@

\section{Results}\label{sec:results}

@@

The primary result is that the \caps{GMM} accurately reflects the performance of the trace data it is based on. It has small, centered error bars for every performance metric presented here, as well as for those we have looked at otherwise. This is an essential result, since otherwise, \caps{GMM} is at best a shaky foundation for further modeling work. With these results, we can be assured that if we can approximate the matrix for a given example of trace traffic, then we have also approximated the original trace behavior. In a sense, this result is a reduction of the general traffic modeling problem to a better-understood matrix modeling problem.

It is worth noting that the \caps{GMM} has competition from a few of the time-simplified models: the \model{time marginal} and \model{time uniform} model both do approximately as well, and in some cases better, for every metric examined. {\FHC} found strong evidence that session arrivals followed a time-varying Poisson arrival process, meaning that other than long-scale time-varying arrival rate and very small-scale clustering of flows within sessions, the overall flow arrival rate is fairly smooth. We suspect that on the scale of minutes to tens of minutes in which our simulation exist, it is sufficient to model both flow and session arrivals uniformly. This result indicates that the precise temporal placement of flows is not highly sensitive: they tend to be relatively evenly spaced out at short time-scales, and network performance is not sensitive to changes in start-time on that scale.

>>>>

After presenting \caps{GMM} and discussing how common concepts in traffic modeling can be expressed in precise mathematical terms within this framework, we proceed with experimental validation of \caps{GMM}. We use the technique of \textit{paired differential simulation} that the authors proposed and used in previous work~\cite{Karpinski07:realism,Karpinski07:cbr-failure}. This technique uses real wireless trace traffic as an experimental control and conducts paired experiments comparing network performance from simulations using control traffic against others using synthetically generated traffic. The results are very positive: \caps{GMM} accurately reproduces network performance characteristics at all levels of the protocol stack.
%This does not solve the ``cookbook'' problem, however, because the \caps{GMM} must still be based on real traffic in order to reproduce its behavioral characteristics. It does however, reduce the problem of general traffic modeling to a more constrained problem of matrix modeling.

%Our experiments with various common assumptions of uniformity and regularity in network behavior yield mixed conclusions: most straightforward assumptions of homogenous behavior distort performance results significantly, but there are notable exceptions. The clearest exception is the time domain: traffic models appear to be relatively insensitive to local perturbation of the start time of flows---so long as the overall distribution remains approximately correct. Our conclusion is that traffic modeling needs to discard its assumptions of uniformity, regularity and independence. Instead we need to move ahead trying to understand the complex dependencies between the many aspects of traffic behaviors. By reducing traffic modeling to the well-understood domain of numerical linear analysis, the General Matrix Model brings us significantly closer to the ultimate goal of having a ``cookbook'' model for generating arbitrary, realistic experimental workload patterns.

@@

The rest of the paper is organized as follows. In \Section{motivation} we discuss motivation and related work. The General Matrix Model is presented in \Section{general-matrix-model}. In Sections~\ref{sec:common-properties}~and~\ref{sec:modeling-concepts} we explore how common network properties and modeling concepts can be expressed in the the \caps{GMM} framework. Our experimental and analytical methodology for evaluating traffic models is presented in \Section{methodology}, while the results of our experiments are explained and analyzed in \Section{results}. The ramifications of these results are discussed in \Section{discussion}. Finally, in \Section{conclusions}, we conclude with a discussion of how this research may be applied to current wireless studies, and how it points the way to better traffic models for the future.

\section{Motivation \& Related Work}
\label{sec:related-work}
\label{sec:motivation}

@@

\section{Expression of Common Properties}
\label{sec:common-properties}

@@

\section{Expression of Common Modeling Concepts}
\label{sec:modeling-concepts}

@@

Since uniform behaviors entail homogeneity of properties with respect to each flow individually, a trace matrix can be transformed to a matrix satisfying the model's assumptions via right matrix multiplication. Right multiplication mixes the elements within each row, and if the mixing weights are uniform, the result is uniform behavior. For example, to make the roles of the nodes with respect to each flow statistically identical, we transform $S$ and $D$:
\begin{align*}
S' &= S\ones{n \times n}/n, & D' &= D\ones{n \times n}/n.
\end{align*}
This makes all the values in each flow's row identically the average of the original $n$ values. As another example, the constant bit-rate (\caps{CBR}) flow model, in its strictest sense, is an intra-flow model, which stipulates that all the packets in each flow behave identically, having identical sizes and intervals; thus each flow's bit-rate is constant, as the name implies. The strict \caps{CBR} model can be easily expressed with the packets-bytes-duration (\caps{PBD}) model used as an example in \Section{representations}. The pair of matrices $\seq{Z,V}$ is transformed to \caps{PBD} form using the vectors defined in \Section{flow-totals}: $\seq{Z,V}\Gamma\in\R^{f \times 3}$, where
\begin{align}
\Gamma=\begin{pmatrix}
\ones{} & \vec{\sigma} & \zeros{} \\
\zeros{} & \zeros{} & \vec{\tau}
\end{pmatrix} \in \R^{(d_v+d_z)\times{3}}.
\end{align}
\Figure{trace-vs-uniform} graphically depicts examples of uniform behaviors as compared to realistic behaviors from traces.
%By replacing $S'$ with $S$, this can also be viewed as a set of linear constraints on the matrix $S$, providing a proscriptive definition

@@

\begin{figure}[b]
%\vspace{-1em}
\begin{center}
\includegraphics[width=3.5in]{nodes-flows}%
\vspace{-0.75em}%
\caption{The number of active nodes and flows over time.} 
\label{fig:nodes-flows}
\end{center}
\vspace{-0.9em}
\end{figure}

We do not assume or claim that the traffic found at \caps{IETF}60 is representative of conference settings in general. The observed behaviors are also unlikely to resemble those found in a typical commercial or residential setting. We have chosen this trace, however, because within it can be found behaviors resembling many different types of wireless usage cases. \Figure{nodes-flows} shows the wide variations in the number of active flows and nodes over the course of the trace. In the night and morning hours, the traffic patterns are similar to those one might find in a moderately trafficked business or residential area. During working group sessions, we see highly concentrated, heavy usage patterns. At the zenith of activity, over 800 users, 33 thousand flows, and 1 million packets are seen in a single 10-minute trace segment. At the nadir, a lone node sent only a single 61-byte packet over the course of 10 minutes. All levels of activity between these extremes are represented. Moreover, the mix of traffic types observed changes dramatically over the course of the day, providing a wide representation of possible blends of behavior. This heterogeneity and extreme range of behaviors makes the \caps{IETF} data set ideal for this evaluation. The variety of activity gives us greater confidence that success or failure of traffic models is not tied to any specific network condition, but is broadly and generally applicable. %For a traffic model to be truly realistic, it must be realistic across many different usage cases.

\break
Before using the traces, it is necessary to extract application-level behavior from the trace header data. First, we split the trace into individual packet flows. A flow is a series of packets sharing the following five attributes: \caps{IP} and transport protocols (raw \caps{IP}, \caps{ICMP}, \caps{TCP}, \caps{UDP}); source and destination \caps{IP} addresses and \caps{TCP}/\caps{UDP} port numbers. Next, the quantity of application-initiated data contained in each packet is calculated. For non-\caps{TCP} packets, this quantity is simply the size of the transport-layer payload, but for \caps{TCP} the calculation is more complicated: only new data transfers, explicitly initiated by the application are counted. Data retransmitted by \caps{TCP} is disregarded, and empty \caps{ACK}s are ignored. \caps{SYN} and \caps{FIN} flags in packets (even empty ones) are counted as a single byte each, since they are explicitly signaled by the application. %After processing, we have a collection of flows, each with a sequence of timestamps and the amount of application-initiated data sent at that time.

@@

We have selected six performance metrics to present here. They are commonly used as indicators of network performance at the application, network, and link layers of the protocol stack:
\begin{enumerate}
\setlength{\itemsep}{0em}
\item \textbf{Application:} average end-to-end delay, packet delivery ratio, received throughput.
\item \textbf{Network:} \caps{AODV} control overhead (\caps{RREQ/RREP/RERR}), packets dropped in routing queues.
\item \textbf{Link:} 802.11 control overhead (\caps{RTS/CTS/ACK}).
\end{enumerate}
These metrics are commonly used to evaluate wireless protocols. We have examined a broad variety of other wireless network performance metrics, and although we do not have room to present or discuss them here, the results shown are representative of the overall realism of the traffic models.
%For each of the 144 10-minute trace segments, we have run simulations using each of the nine traffic models, for a total of 1,296 simulations.
%Unfortunately, not all of the simulations were able to run to completion. In some scenarios, the number of nodes and flows caused Qualnet to simply fail. We were, however, able to get complete results for 72.5\% of the simulations, providing a clear view of overall trends.\!\footnote{That 27.5\% of simulations did not finish may seem like a cause for concern, but it does not in fact invalidate our results. The simulations that did not finish are those for which the number of flows and nodes overwhelmed the capacity of the simulator. Qualnet cannot simulate more than circa 650 nodes sending 35,000 flows. While it would be of great interest to examine results for these large-scale scenarios, they are essentially useless for comparison to other simulations studies: other studies do not simulate this quantity of nodes and flows precisely because even the best existing simulators do not support it. Once the simulators become more scalable and robust, our analysis can be extended to these scenarios as well.}

@@

\section{Results}
\label{sec:results}

@@

The primary result is that the \caps{GMM} accurately reflects the performance of the trace data it is based on. It has small, centered error bars for every performance metric presented here, as well as for those we have looked at otherwise. This is an essential result, since otherwise, \caps{GMM} is at best a shaky foundation for further modeling work. With these results, we can be assured that if we can approximate the matrix for a given example of trace traffic, then we have also approximated the original trace behavior. In a sense, this result is a reduction of the general traffic modeling problem to a more tractable matrix modeling problem.

It is worth noting that \caps{GMM} has competition from a few of the time-simplified models: the \model{time marginal} and \model{time uniform} model both do approximately as well, and in some cases better, for the metrics examined. {\FHC} found strong evidence that session arrivals followed a time-varying Poisson arrival process, meaning that other than long-scale time-varying arrival rate and very small-scale clustering of flows within sessions, the overall flow arrival rate is fairly smooth. We suspect that on the scale of minutes to tens of minutes in which our simulation exist, it is sufficient to model both flow and session arrivals uniformly. This result indicates that the precise temporal placement of flows is not highly sensitive: they tend to be relatively evenly spaced out at short time-scales, and network performance is not sensitive to changes in start-time on that scale.

Not all marginal models have such a benign effect on the accuracy of performance metrics. The worst offender by far is the \model{size marginal} model. This model distorts behavior at every level, by more than a factor 20 in 25\% of scenarios in the case of routing queue packet drops. This is a highly significant result because modeling flow size using a network-wide marginal distribution is precisely what {\FHC}, for example, attempt. This results informs us that this approach is doomed to failure: flow size cannot be assigned without considering its relation to other flow properties: source and destination nodes, and packet behavior at the very least.

The \model{packet marginal} model is another offender, albeit not as egregiously. The direction of the misrepresentations in this case, however, is more dangerous: received throughput is overestimated by more than three times, 75\% of the time. On the other hand, \caps{AODV} control overhead is underestimated by half on average. Such serious misrepresentations are especially noteworthy given the apparently innocuous assumption applied by this model: the \model{packet marginal} model is essentially a standard variable bit-rate (\caps{VBR}) packet behavior model, using the empirical network-wide distributions of packet sizes and inter-packet intervals. Otherwise the model matches the original trace behavior exactly. And yet these drastic distortions occur. In our previous work, we have shown similarly drastic misrepresentations to occur when constant bit-rate packet behavior is assumed. This results indicates that to reproduce accurate network performance, we must provide packet behaviors on a more granular level: each node or even each flow must be assigned a ``custom'' packet behavior by some means.

These strong negative results for two fundamental and very common marginal models provide severe limits on the realism that can be achieved through marginal modeling. Traffic modeling research \textit{must} move beyond simply finding parametric models for marginal distributions of network-wide properties. Realistic behavior cannot be generated using these distributions alone. Something must be captured about the interaction between the elements of flow behavior.
